---
title: "Biostats Wk 3"
author: "MARIAH"
date: "2026-09-12"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

**Research hypothesis:** At least one of the six variables have an effect on wheat yield, and these variables can be used to predict wheat yield.

The response variable is yield (y), it is numerical and continuous. The variables winter plants , mean number of ears, Soil pH, (phosphorus)p, (potassium)k, (magnesium)mg are numerical, contuinous, and focal predictor variables. There are no covariates.

```{r}
library(ggplot2)
wheat <- read.csv("~/Desktop/week 2 bio/Datasets/wheat.csv")
summary(wheat)
```

```{r}
#descriptive stats summary
main<- lm(yield ~ winter+ears+pH+P+K+Mg, data=wheat)
summary(main)
```

## Including Plots
**the residuals are randomly scattered, assumption of homoscedacity is met!**
```{r ggplot2, echo=FALSE}
ggplot(main, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

```{r, echo=FALSE}
print("visual data exploration for all variables")
hist(wheat$yield, main = "Yield", xlab = "Yield")
hist(wheat$winter, main = "Winter plants", xlab = "winter")
hist(wheat$ears, main = "mean number of ears", xlab = "ears")
hist(wheat$pH, main = "Soil pH", xlab = "pH")
hist(wheat$P, main = "Phosphorus", xlab = "P")
hist(wheat$K, main = "Potassium", xlab = "K")
hist(wheat$Mg, main = "Magnesium", xlab = "Mg")
```

```{r, echo=FALSE}
# Pairwise scatterplots for all variable combinations which include the response.
plot(wheat$winter, wheat$yield,
     xlab = "Winter plants",
     ylab = "Yield",
     main = "Yield vs Winter pllants")

plot(wheat$ears, wheat$yield,
     xlab = "Ears ",
     ylab = "Yield",
     main = "Yield vs Mean number of Ears")

plot(wheat$pH, wheat$yield,
     xlab = "Soil pH",
     ylab = "Yield",
     main = "Yield vs Soil pH")

plot(wheat$P, wheat$yield,
     xlab = "Phosphorus",
     ylab = "Yield",
     main = "Yield vs Phosphorus")

plot(wheat$K, wheat$yield,
     xlab = "Potassium",
     ylab = "Yield",
     main = "Yield vs Potassium")

plot(wheat$Mg, wheat$yield,
     xlab = "Magnesium",
     ylab = "Yield",
     main = "Yield vs Magnesium")
```

```{r}
#boxplots DA
boxplot(wheat$winter,
     xlab = "Winter plants",
     ylab = "Yield",
     main = "Yield vs Winter plants",
      col = c("pink"))

boxplot(wheat$ears,
     xlab = "Ears ",
     ylab = "Yield",
     main = "Yield vs Mean number of Ears",
      col = c("purple"))

boxplot(wheat$pH,
     xlab = "Soil pH",
     ylab = "Yield",
     main = "Yield vs Soil pH",
      col = c("blue"))
     
boxplot(wheat$P,
     xlab = "Phosphorus",
     ylab = "Yield",
     main = "Yield vs Phosphorus",
      col = c ("yellowgreen"))

boxplot(wheat$K,
     xlab = "Potassium",
     ylab = "Yield",
     main = "Yield vs Potassium",
      col = c("gold"))

boxplot(wheat$Mg,
     xlab = "Magnesium",
     ylab = "Yield",
     main = "Yield vs Magnesium",
     col = c("tomato"))

```

```{r}
# Scatterplot Matrix
pairs(wheat[, c("yield", "winter", "ears", "pH", "P", "K", "Mg")])
```

```{r}
#correlation matrix
read_w <- cor(wheat[, c("yield", "winter", "ears", "pH", "P", "K", "Mg")], use = "complete.obs")
pairs(read_w)
```

#### *Yield vs pH has a outlier in the lower-left of the plot. pH vs P also has two outliers at the top right.* 

#### *Futhermore, I dont see any clear correlation between any pair of variables.*

Thanks!
